---
title: "Space-Time Prediction of Bike Share Demand: Philadelphia Indego"
author: "Nina Carlson, Ryan Drake, Chloe Robinson, Henry Sywulak-Herr"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

# Executive Summary

Eviction is a pressing issue in the city of Philadelphia. Every year,
thousands of households receive eviction filings. Not only is this
destabilizing for the households involved, but it contributes to
neighborhood instability, weakened social networks, and other
socioeconomic effects that ultimately contribute to inequities. We
produced a model to predict evictions by tract in Philadelphia in order
to help the City mobilize resources and proactively implement targeted
policies that prevent evictions.

Our model uses eviction filing data for Philadelphia at the Census tract
level, paired with crime, affordable housing, housing counseling, market
value analysis, and Census data. We explore where and when evictions
occur to build a svelte predictive model of monthly filings. We
conducted exploratory data analysis after cleaning the data and joining
it to 2020 Census tracts to examine spatial patterns of filings,
temporal trends across the city, and differences across race-majority
tracts. We then integrated multiple data sources to construct a panel
dataset of monthly tract-level predictors from 2022 onward. 

# Methods Overview

We estimated a series of linear regression models that predict monthly
eviction filings. These linear regressions included factors such as
baseline eviction filing (using pre-pandemic averages), demographic
characteristics, crime levels, distance to housing counseling agencies,
neighborhood market value clusters, and recent eviction history
(lagged). Models were trained on data through December 2023 and
evaluated on 2024 onward through MAE.

Generally, we found that adding socio-economic and neighborhood context
modestly improves prediction accuracy. Incorporating recent eviction
lags yields the strongest performance. However, the analysis also
highlights key limitations: under-counting of sealed records,
measurement error across data sources, and ethical concerns about
predicting eviction risk for already-vulnerable communities. 

# Process

## Load Packages

```{r load-packages, message=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman, quietly = T)}
p_load(lubridate, sf, tidycensus, tidygeocoder, tidyverse, tigris, tmap, viridis)

set.seed(5746)
```

## Load Data

### Evictions Dataset

The primary dataset analyzed in this report was derived from the
Evictions Lab's Eviction Tracking System, which has been derived from
county court records for cities such as Philadelphia that have a
reasonably transparent and accessible system of data storage. As a
result, the breadth of records that are often available is limited by
government/judicial cooperation. Furthermore, many city-wide records
were incorrectly geocoded at the time of creation and labeled as
"sealed." Therefore, evictions on average within this dataset are
considered to be an underestimate of the true number of filings in
Philadelphia This data is available at the census tract and zip code
levels for Philadelphia between the years of 2020 and November of 2025
at monthly intervals.

```{r load-data}
# load evictions dataset to investigate
evictions_df <- read_csv("./data/philadelphia_monthly_2020_2021.csv", show_col_types = F)

evictions_df <- evictions_df %>%
  filter(GEOID != "sealed")
```

```{r missing-combos-check}
# generate vectors of all possible months and GEOIDs
months <- seq(as.Date("2020-01-01"),
              as.Date("2025-11-01"),
              by = "month") %>% 
  format("%m/%Y")
geoids  <- unique(evictions_df$GEOID)

# expand a grid of GEOID/month combinations
geoids_months <- expand.grid(GEOID = geoids,
                             month = months,
                             stringsAsFactors = F)

# identify missing combinations
missing <- geoids_months %>% 
  anti_join(evictions_df %>% select(GEOID, month),
            by = c("GEOID", "month"))

# print result
if (nrow(missing) > 0){
  cat("Number of Missing GEOID/Month Combinations:", nrow(missing))
} else{
  cat("No GEOID/Month Combinations Missing")
}
```

The modeling strategy employed in this analysis relies on a panel
dataset which has all combinations of census tract geographic
identification numbers (geoids) and months. As of the 2020 Decennial
Census and the subsequent realignment of census tract boundaries,
Philadelphia County has 408 census tracts, with data spanning 71 months,
resulting in 28,968 records. In anticipation of the availability of
census data, which changed boundaries in 2020 and is only available up
to 2023 (4 years total), the minimum year of evictions data considered
in this analysis was 2022, reducing the overall number of records to
19,176 (-33.8%).

### Census Tracts

```{r load-census-tracts}
# designate FIPS codes for PA and Philly
fips_pa <- 42
fips_phl <- 101

# load tigris census tract boundaries
tracts <- tracts(state = 42, county = 101, year = 2020, progress_bar = F) %>% 
  select(GEOID) %>% 
  st_transform(2272)

# join to evictions dataset and make spatial
evictions <- left_join(evictions_df, tracts, by = "GEOID", ) %>% 
  st_as_sf() %>% 
  mutate(year = year(my(month)),
         year_lag_census = year(my(month)) - 2)
```

The evictions dataset does not include geometries, which instead can be
derived from the Census Bureau's TIGER/line shapefiles. The boundaries
used were derived from the 2020 Decennial Census and joined to based on
geoid.

## Exploratory Data Analysis - Evictions

```{r create-date-format}
# Create proper date format
evictions <- evictions %>%
  mutate(
    month_date = as.Date(paste0("01/", month), format = "%d/%m/%Y"),
    delta = filings_2020 - filings_avg_prepandemic_baseline,
    ratio = filings_2020 / filings_avg_prepandemic_baseline
  )
```

An average filing count calculated for before the COVID-19 pandemic were
used to determine both a delta measure (difference from total filings in
a month) and a ratio of filings in a month relative to this baseline.
This was intended to indicate the relationship of filings at a given
point in time to past trends.

### Total Filings Per Tract

```{r create-total_filings}
# Total filings per tract
tract_totals <- evictions %>%
  group_by(GEOID) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE))

#Plot
ggplot(tract_totals) +
  geom_sf(aes(fill = total_filings), 
          color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "Total Eviction Filings by Census Tract",
    subtitle = "Cumulative filings across all years in the dataset",
    fill = "Total Filings"
  ) +
  theme_void()
```

Total filings per census tract from 2022-2025 are distributed unevenly
throughout Philadelphia, with West Philadelphia and North Philadelphia
seeing far more evictions relative to more affluent areas of the city
such as Center City and South Philadelphia. The greatest number of
filings occurred in the Roxborough Neighborhood adjacent to the
Wissahickon River, as well as on the northern edge of the Northeast
Philadelphia Airport.

### Temporal Trends

#### Total Monthly Filings

```{r create-monthly-evictions-plot}
# Create monthly evictions
eviction_monthly <- evictions %>%
  group_by(month_date) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE))

# Plot total monthly evictions 
ggplot(eviction_monthly, aes(x = month_date, y = total_filings)) +
  geom_line(color = "#3182bd", linewidth = 1) +             # blue line
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +  # red trend line
  labs(
    title = "Monthly Eviction Filings",
    subtitle = "Tract-level eviction filings aggregated across the city",
    x = "Date",
    y = "Total Filings",
    caption = "Source: Eviction Court Filings"
  )
```

Total eviction filings were extremely high at the onset of 2020 as a
consequence of the COVID-19 pandemic, before plummeting sharply in the
middle of the year. They continued to rise into 2023, and have been on a
steady decline since. These rates fluctuate at times significantly from
the trendline over time, but they do not appear to follow a seasonal or
annual pattern.

### Descriptive Distributions

#### Distribution of Monthly Findings

```{r create-monthly-predictions-plot-by-tract}
# Create monthly evictions
# Filings_2020 = total eviction filings
ggplot(evictions %>% filter(filings_2020 <= 50),
       aes(filings_2020)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(title = "Distribution of Monthly Filings per Tract (<= 50)",
       x = "Filings", y = "Count")
```

#### Distribution of Delta

```{r change-from-baseline-distribution}
# Distribution of Delta 
ggplot(evictions, aes(delta)) +
  geom_histogram(bins = 30, fill = "purple") +
  labs(
    title = "Delta: Filings vs Pre-Pandemic Baseline",
    x = "Delta",
    y = "Count"
  )
```

#### Distribution of Ratio

```{r change-from-baseline-distribution-ratio}
# Distribution of Ratio 
ggplot(evictions, aes(ratio)) +
  geom_histogram(bins = 30, fill = "darkgreen") +
  labs(
    title = "Ratio of Filings to Baseline",
    x = "Ratio",
    y = "Count"
  )
```

### Racial Disparaties

```{r mean-filings-by-race}
evictions %>%
  group_by(racial_majority) %>%
  summarise(mean_filings = mean(filings_2020, na.rm = TRUE)) %>%
  ggplot(aes(racial_majority, mean_filings, fill = racial_majority)) +
  geom_col() +
  labs(
    title = "Average Monthly Eviction Filings by Racial-Majority Tract",
    x = "Racial Majority Group",
    y = "Average Monthly Filings"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Filings Over Time by Racial Majority

```{r evictions-by-race}
evictions %>% 
  group_by(month_date, racial_majority) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(month_date, total_filings, color = racial_majority)) +
  geom_line() +
  labs(title = "Eviction Filings Over Time by Racial-Majority Tract",
       x = "Date",
       y = "Total Filings",
       color = "Racial Majority") +
  theme_minimal()
```

Rates of evictions in majority-Black census tracts are consistently
higher than those of majority-White census tracts. While
majority-Hispanic census tracts appear to have lower eviction totals on
average compared to majority-White census tracts, it is important to
note that there are significantly fewer majority-Hispanic census tracts
within the dataset (Hispanic = 5.6%, White = 36.0%, Black = 35.8%, Other
= 22.6%), which obscures potentially high rates of Hispanic household
evictions in census tracts where other race categories dominate.

## Import Predictor Data

### Affordable Housing Production/Availability

Affordable and public housing developments have been found by the
Eviction Lab to be associated with higher-than-normal rates of eviction
filings against tenants, resulting in a pattern of "serial eviction"
([Eviction Lab,
2023](https://evictionlab.org/public-housing-and-the-threat-of-eviction/)).
Affordable housing productions and inventories in Philadelphia for
approximately the past six years are explored below as a potential
predictor.

```{r import-affordable-housing}
housing_sf <- st_read('./data/Affordable_Housing.geojson', quiet = T) %>% 
  st_transform(2272)

# filter out properties for which there is no year associated and geometry is not empty
housing <- housing_sf %>% 
  filter(!is.na(fiscal_year_complete),
         !st_is_empty(.)) %>% 
  st_join(tracts, join = st_within)
```

```{r message=FALSE}

housing_cnts <- housing %>% 
  st_drop_geometry() %>% 
  group_by(GEOID, fiscal_year_complete) %>% 
  summarise(n_proj = n(),
            total_units = sum(total_units)) %>%
  mutate(n_proj_cum = ave(n_proj, FUN = cumsum),
         total_units_cum = ave(total_units, FUN = cumsum)) %>% 
  filter(fiscal_year_complete >= 2020) %>% 
  left_join(tracts, by = "GEOID") %>% 
  st_as_sf(crs = 2272)
```

```{r viz-affordable-housing}
ggplot() +
  geom_sf(data = tracts,
          fill = NA,
          color = 'grey70') +
  geom_sf(data = housing, 
          mapping = aes(size = total_units),
          lwd = 0,
          color = 'purple4',
          alpha = 0.25) +
  labs(title = "Affordable Housing in Philadelphia",
       subtitle = "Locations and Total Unit Counts",
       size = 'Total Units') +
  theme_void()
```

Affordable housing locations in Philadelphia are spatially concentrated
immediately north and south of Center City as well as Northwest
Philadelphia. Notable areas devoid of affordable housing developments
are Northeast Philadelphia and

```{r}
ggplot() +
  geom_sf(data = tracts, color = 'grey85', fill = NA) +
  geom_sf(data = housing_cnts, aes(fill = total_units_cum)) + 
  facet_wrap(~fiscal_year_complete) +
  theme_void()
```

### Crime

```{r load-crime}
# Path to the main directory that contains the crimeXXXX folders
crime_main_dir <- "./data/crime/"

# 1. Find all subdirectories matching "crimeXXXX"
crime_subdirs <- list.dirs(crime_main_dir, full.names = TRUE, recursive = FALSE)

# 2. Find all .shp files inside those directories
crime_files <- list.files(crime_subdirs, pattern = "\\.shp$", full.names = TRUE)

# 3. Read all shapefiles into a list
crime_list <- lapply(crime_files, function(x) {
  st_read(x, quiet = T) %>% 
    st_transform(2272)
  }) %>% do.call(rbind, .)

crime <- crime_list %>% filter(!if_any(c('point_x', 'point_y'), is.na))
```

```{r message=FALSE}
# join tract geoid to crimes and filter out those not in Philly
crime_tract <- st_join(crime, tracts, join = st_intersects) %>% 
  filter(!is.na(GEOID))

# create a month/year column
crime_tract <- crime_tract %>% 
  mutate(month = format(dispatch_d, format = "%m/%Y"))

# summarize by the number of each type of crime per census tract per month
crime_cnts <- crime_tract %>% 
  st_drop_geometry() %>% 
  group_by(GEOID, month) %>% 
  summarise(count = n()) %>% 
  ungroup() %>%  
  complete(data = ., GEOID, month, fill = list(count = 0))
```

#### Crime EDA

```{r message=FALSE}
# Fix date format
crime_cnts <- crime_cnts %>% 
  mutate(
    month_date = as.Date(paste0("01/", month), format = "%d/%m/%Y"),
    year = lubridate::year(month_date)
  )

# Crimes Over Time in Philadelphia
crime_cnts %>% 
  group_by(month_date) %>% 
  summarise(total_crime = sum(count)) %>% 
  ggplot(aes(month_date, total_crime)) +
  geom_line(color = "firebrick") +
  labs(
    title = "Total Crime Over Time (Citywide)",
    x = "Month",
    y = "Incidents"
  ) +
  theme_minimal()
```

```{r message=FALSE}
crime_totals <- crime_cnts %>% 
  group_by(GEOID) %>% 
  summarise(total_crime = sum(count))

ggplot(crime_totals, aes(total_crime)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Total Crime Counts Across Census Tracts",
    x = "Total Crime (2022–2025)",
    y = "Number of Tracts"
  ) +
  theme_minimal()
```

```{r message=FALSE}
#Yearly Crime
crime_yearly <- crime_cnts %>% 
  group_by(GEOID, year) %>% 
  summarise(total_crime = sum(count), .groups = "drop") %>% 
  left_join(tracts, by = "GEOID") %>% 
  st_as_sf()

ggplot(crime_yearly) +
  geom_sf(aes(fill = total_crime),
          color = "grey80",
          linewidth = 0.1) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(na.value = "grey95") +
  labs(
    title = "Crime Counts by Census Tract (Annual)",
    subtitle = "Exploratory spatial analysis",
    fill = "Crime count"
  ) +
  theme_void()

# Census tracts with the most crime
crime_totals_ranked <- crime_cnts %>% 
  group_by(GEOID) %>% 
  summarise(total_crime = sum(count)) %>% 
  arrange(desc(total_crime))

head(crime_totals_ranked, 10)

# Total crime counts per census tract from 2022-2025
crime_map <- tracts %>% left_join(crime_totals_ranked, by = "GEOID")

ggplot(crime_map) +
  geom_sf(aes(fill = total_crime),
          color = "grey80",
          linewidth = 0.1) +
  scale_fill_viridis_c(na.value = "grey95") +
  labs(
    title = "Total Crime by Census Tract",
    subtitle = "2022-2025",
    fill = "Crime count"
  ) +
  theme_void()
```

### Market Value Analysis

```{r}
# read in results of 
mva <- st_read('./data/mva_2023.geojson', quiet = T) %>%
  st_transform(2272)

# the data is at the block group level
# GEOID (12 digits) = tract (11 digits) + block grp (1 digit)
mva <- mva %>% 
  mutate(GEOID = substr(geoid, 1, 11),
         clusterletter = ifelse(clusterletter == " ", NA, clusterletter))

# determine the dominant 
mva_ranks <- mva %>% 
  st_drop_geometry() %>% 
  group_by(GEOID) %>% 
  summarise(
    rank = {
      x <- na.omit(clusterletter)     # remove NA
      if (length(x) == 0) {
        NA_character_                # if no letters, return NA
      } else {
        tbl <- table(x)
        max_letters <- names(tbl)[tbl == max(tbl)]
        sort(max_letters)[1]
      }
    },
    .groups = "drop"
  )
```

### Market Value Analysis - EDA

```{r}
mva_ranks %>% 
  summarise(
    n_tracts = n_distinct(GEOID),
    n_missing = sum(is.na(rank)),
    n_categories = n_distinct(rank, na.rm = TRUE)
  )

table(mva_ranks$rank)

prop.table(table(mva_ranks$rank))

# Most common MVA categories
ggplot(mva_ranks, aes(rank)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of MVA Cluster Categories",
    x = "MVA Cluster",
    y = "Number of Census Tracts"
  ) +
  theme_minimal()
```

```{r}
# MVA dominant cluster by tract
mva_ranks_sf <- tracts %>% 
  left_join(mva_ranks, by = "GEOID")

ggplot(mva_ranks_sf) +
  geom_sf(aes(fill = rank), 
          color = "grey80",
          linewidth = 0.1) +
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "grey90") +
  labs(
    title = "Market Value Analysis (MVA) — Dominant Cluster by Tract",
    subtitle = "Exploratory spatial visualization",
    fill = "Cluster"
  ) +
  theme_void()
```

## Housing Counseling Agencies

```{r}
# create k-nearest neighbors function
get_knn_distance <- function(dist_matrix, k) {
  apply(dist_matrix, 1, function(distances) {
    mean(as.numeric(sort(distances)[1:k]))
  })
}
```

```{r warning=FALSE}
# load housing counseling agency locations
hca <- st_read('./data/HousingCounselingAgencies.geojson', quiet = T) %>% 
  st_transform(2272)

# create centroids for each tract to calculate distances
tracts_cent <- tracts %>% 
  st_centroid()

# claculate distance matrix
hca_distmatrix <- st_distance(tracts_cent, hca)

# distance to the nearest 1 and nearest 3 HCAs
hca_dist <- tracts %>% 
  mutate(dist_1hca = get_knn_distance(hca_distmatrix, k = 1),
         dist_3hca = get_knn_distance(hca_distmatrix, k = 3))
```

### Housing Counseling Agencies - EDA

```{r warning=FALSE}
# Histogram Distance to nearest HCA
ggplot(hca_dist, aes(dist_1hca)) +
  geom_histogram(bins = 40, fill = "purple", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Distance to Nearest HCA",
    x = "Distance (meters)",
    y = "Number of Census Tracts"
  )

# Histogram Distance to nearest 3 HCA's
ggplot(hca_dist, aes(dist_3hca)) +
  geom_histogram(bins = 40, fill = "darkgreen", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Distance to 3 Nearest HCAs",
    x = "Distance (meters)",
    y = "Number of Census Tracts"
  )

#Scatterplot dist_1 vs. dist_2
ggplot(hca_dist, aes(dist_1hca, dist_3hca)) +
  geom_point(alpha = 0.6, color = "purple") +
  theme_minimal() +
  labs(
    title = "Relationship Between 1-HCA and 3-HCA Distances",
    x = "Distance to nearest HCA",
    y = "Mean distance to 3 nearest HCAs"
  )

```

```{r warning=FALSE}
# Map dist to nearest HCA
ggplot(hca_dist) +
  geom_sf(aes(fill = dist_1hca), color = "grey80") +
  scale_fill_viridis_c() +
  theme_void() +
  labs(
    title = "Distance to Nearest Housing Counseling Agency",
    fill = "Meters"
  )

# Map dist to nearest 3 HCA
ggplot(hca_dist) +
  geom_sf(aes(fill = dist_3hca), color = "grey80") +
  scale_fill_viridis_c() +
  theme_void() +
  labs(
    title = "Distance to 3 Nearest Housing Counseling Agencies",
    fill = "Meters"
  )

```

## Census Demographic Variables

```{r message=FALSE}
# create a vector of years to pull from the census
years <- c(2020:2023)

# save vars and their column names to pull
vars <- c('pop_tot' = 'B01003_001',   # total tract population
          'pop_white' = 'B02001_002', # white population (for %-non-white calc)
          'units_tot' = 'B25001_001', # total housing units
          'hh_tot' = 'B11001_001',    # total households
          'hh_fam' = 'B11001_002')    # family households)

# pull demographic variables for all specified years and add a year column
demo_vars <- map_df(years, function(y) {
  get_acs(
    geography = "tract",
    variables = vars,
    year = y,
    survey = "acs5",
    output = 'wide',
    state = fips_pa,
    county = fips_phl,
    progress_bar = F
  ) %>%
  mutate(year_lag_census = y)
})

```

```{r combine-vars}
# isolate evictions in 2022 or later, removing unnecessary columns
evictions_trim <- evictions %>% filter(my(month) >= '2022-01-01') %>% 
  select(-c(type, last_updated))

data_1 <- left_join(evictions_trim, crime_cnts, by = c('GEOID', 'month'))
data_2 <- left_join(data_1, hca_dist %>% st_drop_geometry(), by = 'GEOID')
data_3 <- left_join(data_2, mva_ranks, by = 'GEOID')
data_model <- left_join(data_3, demo_vars,   by = c("GEOID", "year_lag_census")
)
```

#Predictive Modelling

Building complexity over multiple models

#create lag variables

```{r lag and rolling average prep}

data_model <- data_model %>%
  arrange(GEOID, month_date) %>%
  group_by(GEOID) %>%
  mutate(
    # one month and two month lags
    lag_1 = lag(filings_2020, 1),
    lag_2 = lag(filings_2020, 2),

  ) %>%
  ungroup()



```

#Split Train/Test

```{r prep for modelling}

# split date
split_date <- as.Date("2024-01-01")

# Ensure month_date is a proper date
data_model <- data_model %>%
  mutate(month_date = as.Date(month_date))

# Temporal split
train_data <- data_model %>%
  filter(month_date < split_date)

test_data <- data_model %>%
  filter(month_date >= split_date)

# Check split sizes and ranges
nrow(train_data)
nrow(test_data)

summary(train_data$month_date)
summary(test_data$month_date)


```

#Model 0 Baseline

```{r add variables}

model_0 <- lm(
  filings_2020 ~  filings_avg + filings_avg_prepandemic_baseline,
  data = train_data
)

summary(model_0)

```

#Model 1 Demographics

```{r demographics }

# Model 1: baseline + ACS demographic predictors
train_data <- train_data %>%
  mutate(
    pct_nonwhite = 1 - (pop_whiteE / pop_totE),
    hh_fam_share = hh_famE / hh_totE,
    units_per_hh = units_totE / hh_totE
  )

test_data <- test_data %>%
  mutate(
    pct_nonwhite = 1 - (pop_whiteE / pop_totE),
    hh_fam_share = hh_famE / hh_totE,
    units_per_hh = units_totE / hh_totE
  )

```

```{r Model 1}
model_1 <- lm(
  filings_2020 ~ 
    filings_avg_prepandemic_baseline +  # baseline anchor
    pop_totE +
    pct_nonwhite +
    hh_totE +
    hh_fam_share +
    units_totE +
    units_per_hh,
  data = train_data
)

summary(model_1)


```

#Model 2 Crime

```{r model prep crime}

train_data <- train_data %>%
  mutate(crime_cnt = as.numeric(count))

test_data <- test_data %>%
  mutate(crime_cnt = as.numeric(count))


```

```{r crime}

model_2 <- lm(
  filings_2020 ~ 
    filings_avg_prepandemic_baseline + 
    pop_totE +
    pct_nonwhite +
    hh_totE +
    hh_fam_share +
    units_totE +
    units_per_hh +
    crime_cnt,
  data = train_data
)

summary(model_2)


```

#Model 3 HCA Distances

```{r HCA prep}
train_data <- train_data %>%
  mutate(
    dist_1hca = as.numeric(dist_1hca),
    dist_3hca = as.numeric(dist_3hca)
  )

test_data <- test_data %>%
  mutate(
    dist_1hca = as.numeric(dist_1hca),
    dist_3hca = as.numeric(dist_3hca)
  )


```

```{r HCA}

model_3 <- lm(
  filings_2020 ~ 
    filings_avg_prepandemic_baseline + 
    pop_totE +
    pct_nonwhite +
    hh_totE +
    hh_fam_share +
    units_totE +
    units_per_hh +
    crime_cnt +       
    dist_1hca +         
    dist_3hca,          
  data = train_data
)

summary(model_3)

```

#Model 4 Affordable Housing and mva

#still need to add count to affordable housing but need to change name
of column

```{r mva prep}
train_data <- train_data %>%
  mutate(rank = factor(rank))

test_data <- test_data %>%
  mutate(rank = factor(rank))

test_data$rank <- factor(test_data$rank, levels = levels(train_data$rank))

```

```{r affordable housing and mva}
model_4 <- lm(
  filings_2020 ~ 
    filings_avg_prepandemic_baseline + 
    pop_totE +
    pct_nonwhite +
    hh_totE +
    hh_fam_share +
    units_totE +
    units_per_hh +
    crime_cnt +
    dist_1hca +
    dist_3hca +
    rank,                
  data = train_data
)

summary(model_4)

```

#Model 6 Eviction Lags

```{r temporal variables}

model_5 <- lm(
  filings_2020 ~ 
    filings_avg_prepandemic_baseline +
    pop_totE + pct_nonwhite + hh_totE + hh_fam_share +
    units_totE + units_per_hh +
    crime_cnt +
    dist_1hca + dist_3hca +
    rank +
    lag_1 +
    lag_2,
  data = train_data
)

summary(model_5)


```

# Calculate Predictions and MAE

```{r calculate_mae_evictions}
# Generate predictions
test_data <- test_data %>%
  mutate(
    pred_0 = predict(model_0, newdata = .),
    pred_1 = predict(model_1, newdata = .),
    pred_2 = predict(model_2, newdata = .),
    pred_3 = predict(model_3, newdata = .),
    pred_4 = predict(model_4, newdata = .),
    pred_5 = predict(model_5, newdata = .)
  )

# Build MAE results table
mae_results <- tibble::tibble(
  Model = c(
    "0. Baseline",
    "1. + Demographics",
    "2. + Crime",
    "3. + HCA Distance",
    "4. + MVA Rank",
    "5. + Lags"
  ),
  MAE = c(
    mean(abs(test_data$filings_2020 - test_data$pred_0), na.rm = TRUE),
    mean(abs(test_data$filings_2020 - test_data$pred_1), na.rm = TRUE),
    mean(abs(test_data$filings_2020 - test_data$pred_2), na.rm = TRUE),
    mean(abs(test_data$filings_2020 - test_data$pred_3), na.rm = TRUE),
    mean(abs(test_data$filings_2020 - test_data$pred_4), na.rm = TRUE),
    mean(abs(test_data$filings_2020 - test_data$pred_5), na.rm = TRUE)
  )
)

# Show table
knitr::kable(
  mae_results,
  digits = 3,
  caption = "Mean Absolute Error by Model (Test Set)",
  col.names = c("Model", "MAE (filings)")
)
```

# Visualize Model Comparison

```{r visualize model comparison}

mae_results <- mae_results %>%
  mutate(Model = factor(Model, levels = Model))

ggplot(mae_results, aes(x = Model, y = MAE)) +
  geom_col() +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.3, size = 3) +
  labs(
    title = "Test-Set MAE by Model",
    x = "Model",
    y = "MAE (filings)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

#Spatial Errors

TBD

#Temporal Errors

TBD
